DRAFT: Harmonized and trait aggregation paper

### 2020-12-08 Paper/Draft_harmonization_paper.tex
08.12: Update Discussion/Conclusion

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 1. Trait Database 
%- Unterschiede zwischen Regionen und Trait Genauigkeit
%- Was kann man machen
%- Andere Studien zur Trait Harmonisierung usw.

% Why do we need trait-based analysis, especially across regions?
Synthesizing trait information from multiple trait databases is crucial to unfold the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt of harmonising invertebrate trait information from trait databases of different regions showed that grouping feature harmonisation is labour-intensive because grouping features are differentiated differently into traits between trait databases and various codings are used to describe traits. In addition, the same traits are sometimes defined differently in the investigated databases, requiring expert knowledge to reclassify these traits (e.g. the trait piercer in the Tachet database, Table \ref{tab:traits_harmonization}). Although, others have noted the lack of standardized trait terminology in freshwater ecology \cite{baird_toward_2011, brink_traits-based_2011} we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features. To resolve definition discrepancies and facilitate data synthesis terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past, e.g. for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative \cite{garnier_towards_2017}. The core of this harmonisation initiative is to provide standardized trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater science to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could then be linked through standardised terminologies or ontologies, as suggested by Baird et al. \cite{baird_toward_2011} almost a decade ago. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS) providers of invertebrate trait data could connect their traits to such ontologies \cite{schneider_towards_2019}. Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately improve interpretation of derived results. % What is not mentioned here is how traits are measured, for which measurement protocols exist I believe. I think this is an separate issue.

% What's the problem with the state of data coverage
Our analysis aimed to use most of the available invertebrate trait information for different regions to establish harmonised grouping feature datasets. Although many taxa are covered in the trait databases from which we extracted our trait datasets, the availability of trait information varies strongly across grouping features and taxonomic resolution differs between databases. While trait information from Europe and New Zealand was mainly on species-level, considerable portions of trait data from North America and Australia were on genus or family-level (Table \ref{tab:tax_coverage}). Assigned traits on family-level might not reflect the real trait diversity within a taxon, for example for Chironomidae \cite{serra_synthesising_2016}. However, species identification can be complicated and time-consuming \cite{marshall_taxonomic_2006, resh_which_2008}, and therefore trait information on family-level has been widley used e.g. in bioassessment \cite{beketov_spear_2009}.
% How would obtaining information on species level change our results?

@@

@@

In the case when trait values are only available at species-level and observational data are on less precise taxonomic levels various trait aggregation methods have been used by researchers. We aggregated traits from an Australian and North American trait dataset to family-level using 5 different aggregation methods and compared the results to traits assigned at family-level for these regions. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular family is. Aggregation of trait information on species or genus-level to a point estimate on family-level suggests a precision which is not necessarily present, especially not for traits with a high variability or if trait information on species-level is missing. Some traits can vary strongly on more precise taxonomic levels than the family-level. For example, Monaghan and Soares found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction \cite{monaghan_improving_2013}. Furthermore, studies focusing on the lability of traits, i.e. how much traits are unconstrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable \cite{poff_functional_2006, wilkes_traitbased_2020}. However, if trait aggregation is necessary our study can give guidance, under which circumstances the choice of aggregation method might be important. The 5 aggregation methods we tested use either the mean or the median and different weightings. Our results indicated that 1) aggregation of traits using \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{mean}} (median aggregation methods) compared to \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_mean} (mean aggregation methods) is often closer to the assigned traits at family-level and 2) that the different weightings tested exerted a minor impact on the aggregated trait affinities. We obtained for both datasets more differing cases for mean aggregation methods, irrespective of the weighting. In the Australian dataset, the amount of differing cases was slightly higher for the mean aggregation methods (23 \%) than for median aggregation methods (16 to 18 \% differing cases). In the North American dataset, mean aggregation methods yielded in 47 \% of the cases a difference to the assigned traits, much higher in comparison to the median aggregation methods (between 15 and 18 \%, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). However, when there were differences between median aggregation methods and assigned traits, these differences were greater compared to the mean aggregation methods, particularly for the North American trait dataset. The mean absolute differences for the median aggregations was twice as high (0.63 - 0.7) compared to the mean aggregations (0.3 - 0.31, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). The differences can be explained by the binary coding used in the North American trait dataset and in the assigned traits for North America. As a result, traits have affinities of either 1 or 0. Therefore, traits of a particular grouping feature have in most cases a higher difference in trait affinities to each other than if they were fuzzy coded. Using the median to aggregate binary coded taxa will often result in a value of 0 or 1, while using the mean will often result in values between 0 and 1. Thus, using the median led for the North American dataset in many cases either to agreement with the assigned traits or the difference between median aggregated traits and assigned traits was often 1. 

@@

% Binary coding also explains the amount of differing cases, median often right for binary data, must be cautious when a lot of traits are unknown?
The number of differing cases and the mean absolute differences to the assigned traits were similar within the mean aggregation and within the median aggregation methods, which confirms our second finding, the small influence of weighting on the aggregation methods. Furthermore, the distributions of absolute trait affinity differences to assigned traits obtained were relatively similar for the mean aggregation methods and the median aggregation methods in both tested datasets respectively (Figure \ref{fig:diff_aggr_traits_chessman} and Figure \ref{fig:diff_aggr_traits_pyne}). The minor impact on trait aggregation of the weightings may be explained by the fact that a considerable portion of taxa had low numbers of genera or species entries. From the taxa that were compared from the North American trait dataset to the assigned traits at family-level, 14 \% were resolved on family-level and 62 \% were resolved on genus-level. 52 \% had 5 or less genera and 13 \% contained just one genus (Figure \ref{fig:tax_hierarchy_NOA}). In the Australian dataset 21 \% of the taxa were resolved on family-level, 40 \% were resolved on genus level, and 39 \% were resolved on species-level. 68 \% of the taxa contain 5 or less genera and 40 \% just contained one genus (Figure \ref{fig:tax_hierarchy_AUS}).

Although the simulation results showed only small differences between the results of the aggregation methods the choice of the aggregation method for small numbers of genera per family can be important when one genus has a much larger number of species compared to other genera within a particular family. 
% R: Das Argument ist schwer greifbar für den Leser - es wäre einfacher zu folgen, wenn Du schreiben würdest, wann man denn mit unterschieden rechnen würde und warum - und dann deine Daten beschreiben. Also wenn es sehr viele Genera und diese mit sehr unterschiedlicher Anzahl an Arten wären usw. . Die Diskussion sollte vor allen Dingen allgemeiner verständniclih sein und nicht technisch die Resultate fortsetzen.

% R: Misses a starting sentence, Why did we conduct this analysis?
% Mention that comparison to other studies not possible since they don't exist?
In the re-analysis where we investigated how using harmonised and aggregated traits might change the results of trait-environment relationships we found only slightly different results compared to the original study. Some of the traits that responded in the original analysis did not respond in the re-analysis. Those traits that responded only in the original analysis but were non-response in the re-analysis were those closest to the criterion that defined when a trait was associated with either high or low salinity (mahalanobis distance greater than the 97.5 \% quantile of the Chi-square distribution, see methods), namely feeding mode shredder, respiration gills, and life cycle duration traits. A reason for their non-responsiveness was likely the harmonisation procedure. For example, during harmonisation the trait shredder has been amalgamated based on three traits (Miner, xylophagus, and shredder). As a result, the trait affinities in the original dataset have a higher mean and standard deviation than in the aggregated and harmonized datasets (Table S\ref{tab:SI_resp_traits_summary_stats}). Hence, the signal in the original data for the trait shredder has presumably been weakened by the harmonisation, which might be the reason that shredders showed a less strong reaction to the salinity gradient in the re-analysis. The fact that the family-level aggregated traits showed similar results compared to the original analysis is in line with the findings of others that family-level traits can be sensitive enough to detect environmental impacts \cite{beketov_spear_2009}, although species-level traits are expected to result in more precise results when investigating trait-environment relationships. % Cite Schmidt Kloiber? Cite another paper that uses family-level trait data

@@

% What do we really know because someone has watched the animals and published their behaviour; what is extrapolated knowledge? 
% Quality evaluation of traits als Thema..
% What do we learned: 
Comprehensive invertebrate trait databases have been developed over the last years in freshwater ecology. We showed that trait definitions are distinct across these databases. By providing an overview of definition discrepancies we hope to provide a starting point for the development of standardized trait definitions through which trait databases can be linked. Standardized terminolgy will facilitate comparable trait analysis across large scales. 
% - Variation in taxonomic coverage 
If trait aggregation is necessary we showed that using the median is often in conclusion with expert assignments. 
% Different weighting approaches did
Which method is objectively the a better measure cannot be answered, but traits with high variance should be treated carefully and taxa where a few genera have a disproportional high number of species trait information. If researchers need to aggregate traits, a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate.

% Efforts in creating trait ontologies etc etc
% Technology improvement, e.g. Deep learning for species identification

>>>>

% 1. Trait Database 
%- Unterschiede zwischen Regionen und Trait Genauigkeit
%- Was kann man machen
%- Andere Studien zur Trait Harmonisierung usw.

\subsection*{Trait definition discrepancies and taxonomic resolution}

% Why do we need trait-based analysis, especially across regions?
Synthesizing trait information from multiple trait databases is crucial to unfold the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt of harmonising invertebrate trait information from trait databases of different regions showed that grouping feature harmonisation is labour-intensive because grouping features are differentiated differently into traits between trait databases and various codings are used to describe traits. In addition, the same traits are sometimes defined differently in the investigated databases, requiring expert knowledge to reclassify these traits (e.g. the trait piercer in the Tachet database, Table \ref{tab:traits_harmonization}). Although, others have noted the lack of standardized trait terminology in freshwater ecology \cite{baird_toward_2011, brink_traits-based_2011} we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features. To resolve definition discrepancies and facilitate data synthesis in the future terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past for other organisms, e.g. for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative \cite{garnier_towards_2017}. The core of this harmonisation initiative is to provide standardized trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater ecologists to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could be linked through standardised terminologies or ontologies, as suggested by Baird et al. \cite{baird_toward_2011} almost a decade ago. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS) providers of invertebrate trait data could connect their traits to such ontologies \cite{schneider_towards_2019}. Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately interpretation of the derived results. % What is not mentioned here is how traits are measured, for which measurement protocols exist I believe. I think this is an separate issue.

% ?What's the problem with the state of data coverage
Our analysis aimed to use most of the available invertebrate trait information for different regions to establish harmonised grouping feature datasets. Although many taxa are covered in the trait databases from which we extracted our trait datasets, the availability of trait information varies strongly across grouping features and taxonomic resolution differs between databases. While trait information from Europe and New Zealand was mainly on species-level, considerable portions of trait data from North America and Australia were on genus or family-level (Table \ref{tab:tax_coverage}). Assigned traits on family-level might not reflect the real trait diversity within a taxon, for example for Chironomidae \cite{serra_synthesising_2016}. However, species identification can be complicated and time-consuming \cite{marshall_taxonomic_2006, resh_which_2008}, and therefore trait information on family-level has been widely used, e.g. in bioassessment \cite{beketov_spear_2009}.
% How would obtaining information on species level change our results?

@@

\subsection*{Trait aggregation and grouping feature harmonisation}

@@

In the case when trait values are only available at species-level and observational data are on less precise taxonomic levels various trait aggregation methods have been used by researchers. We aggregated traits from an Australian and North American trait dataset to family-level using 5 different aggregation methods and compared the results to traits assigned at family-level for these regions. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular family is. Aggregation of trait information on species or genus-level to a point estimate on family-level suggests a precision which is not necessarily present, especially not for traits with a high variability or if trait information on species-level is missing. Some traits can vary strongly on more precise taxonomic levels than the family-level. For example, Monaghan and Soares \cite{monaghan_improving_2013} found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction. Furthermore, studies focusing on the lability of traits, i.e. how much traits are unconstrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable \cite{poff_functional_2006, wilkes_traitbased_2020}. However, if trait aggregation is necessary our study can give guidance, under which circumstances the choice of aggregation method might be important. The 5 aggregation methods we tested use either the mean or the median and different weightings. Our results indicated that 1) the aggregation methods \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{mean}} (median aggregation methods) compared to \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_mean} (mean aggregation methods) were often closer to the assigned traits at family-level and 2) that the different weightings tested exerted a minor impact on the aggregated trait affinities. We obtained for both datasets more differing cases for mean aggregation methods, irrespective of the weighting. In the Australian dataset, the amount of differing cases was slightly higher for the mean aggregation methods (23 \%) than for median aggregation methods (16 to 18 \% differing cases). In the North American dataset, mean aggregation methods yielded in 47 \% of the cases a difference to the assigned traits, much higher in comparison to the median aggregation methods (between 15 and 18 \%, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). When differences occurred between median aggregation methods and assigned traits, these differences were greater compared to the mean aggregation methods, particularly for the North American dataset. The mean absolute differences for the median aggregations was twice as high (0.63 - 0.7) compared to the mean aggregations (0.3 - 0.31, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). The differences can be explained by the binary coding used in the North American trait dataset and in the assigned traits. As a result, traits have affinities of either 1 or 0. Therefore, traits of a particular grouping feature have in most cases a higher difference in trait affinities to each other than if they were fuzzy coded. Using the median to aggregate binary coded taxa will often result in a value of 0 or 1, while using the mean will often result in values between 0 and 1. 
% Was folgt daraus?
%Thus, using the median led for the North American dataset in many cases to agreement with the assigned traits. However, when differences were found the difference between median aggregated traits and assigned traits was often 1. 

@@

The weightings explored in this analysis should lead to different aggregated trait affinities in cases where several genera within a family have different numbers of species. In fact, the simulation results indicated that for a particular family where one genus has a much larger number of species compared to other genera within this family , especially for traits with a higher variance. However, in the comparison of the trait aggregation methods to assigned traits the number of differing cases and the mean absolute differences in trait affinities were similar across the mean aggregation and across the median aggregation methods, which suggests a small influence of the tested weightings on the aggregation methods. Also, the distributions of absolute trait affinity differences to assigned traits were relatively similar for the mean aggregation methods and the median aggregation methods respectively (Figure \ref{fig:diff_aggr_traits_chessman} and Figure \ref{fig:diff_aggr_traits_pyne}). The minor impact on trait aggregation of the weightings may be explained by the fact that a considerable portion of taxa had low numbers of genera or species entries. From the taxa that were compared from the North American trait dataset, 14 \% were resolved on family-level and 62 \% were resolved on genus-level. 52 \% had 5 or less genera and 13 \% contained just one genus (Figure \ref{fig:tax_hierarchy_NOA}). In the Australian dataset, 21 \% of the taxa were resolved on family-level, 40 \% were resolved on genus level, and 39 \% were resolved on species-level. 68 \% of the taxa contain 5 or less genera and 40 \% just contained one genus (Figure \ref{fig:tax_hierarchy_AUS}). 
% Results might change when trait information is all on species level

In our second analysis we explored how using harmonised grouping features and aggregated traits might change the results of trait-environment relationships. We are not aware of other studies that compared trait-environment relationships using harmonised and not harmonised data. The re-analysis yielded only slightly different results compared to the original study, i.e. some of the traits that responded in the original analysis did not respond in the re-analysis. Those traits that were non-responsive in the re-analysis were those closest to the criterion that defined when a trait was associated with either high or low salinity (mahalanobis distance greater than the 97.5 \% quantile of the Chi-square distribution, see method section), namely feeding mode shredder, respiration gills, and life cycle duration traits. Feeding mode shredder was the only trait that was non-responsive when using harmonised (but not aggregated) as well as when using aggregated data. One reason for the non-responsiveness of shredders in the re-analysis was likely the harmonisation procedure. For example, during harmonisation the trait shredder has been amalgamated based on three traits (Miner, xylophagus, and shredder). As a result, the trait affinities in the original data have a higher mean and standard deviation than in the aggregated and harmonized data (Table \ref{tab:SI_resp_traits_summary_stats}) suggesting that the signal in the original data has presumably been weakened by the harmonisation. This might be the reason that shredders showed a less strong reaction to the salinity gradient in the re-analysis. These findings show that if harmonisation is necessary, harmonised and non-harmonised data should be compared and possible averaging effects should be considered in further analysis.

The fact that the family-level aggregated traits used in the re-analysis showed only slightly different results compared to the original analysis is in line with the findings of others that family-level traits can be sensitive enough to detect environmental impacts \cite{beketov_spear_2009}.
% Although species-level traits are expected to result in more precise results when investigating trait-environment relationships. % Cite Schmidt Kloiber? Cite another paper that uses family-level trait data

@@

% What do we really know because someone has watched the animals and published their behaviour; what is extrapolated knowledge? 
% Quality evaluation of traits als Thema..
Although comprehensive invertebrate trait databases have been developed for several regions in the past, data synthesis is difficult due to discrepancies in trait definitions. By providing an overview of definition discrepancies we hope to set a starting point for the development of standardized trait definitions through which trait databases can be linked. Standardized terminology will facilitate comparable trait analysis across large scales. 
% - Variation in taxonomic coverage 
We showed that trait aggregation is often in conclusion with expert assignments and specifically approaches using the median. ... weighting trait information based on the taxa present in the trait databases exerts a minor influence on the aggregated trait affinities. Which method the objectively better measure is cannot be answered, but when trait aggregation is necessary we showed 1) that highly variable traits should be treated carefully, 2) coding of traits should be considered, and 3) taxa where a few genera have a disproportional high number of species trait information .... If researchers need to aggregate traits, a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate. % Question for future research?
% Fuzzy coding and uncertainty?
% Efforts in creating trait ontologies etc etc
% Technology improvement, e.g. Deep learning for species identification
